

# Introduction

This document provides R-codes that demonstrate the usage of
LTMLE.

Note that the data for the exercises are computer simulated and not
real. The aim is to illustrate the usage of the software `rtmle` for
longitudinal causal inference. 


# Code for practicals III


## Read the data

Load the R-packages and read in the data. They are given in a format
which correspond to electronic health records (register data) with
dates in calendar time.

```{r   }
library(data.table)
library(rtmle)
library(glmnet)
url <- "https://raw.githubusercontent.com/tagteam/doctoral-course-longitudinal-causal-inference/main/practicals/data/register_data.RDS"
register_data <- readRDS(url(url))
str(register_data) 
```

The data contain baseline information:

-   `gender` person gender
-   `bslkarnofsky` Karnofsky score at time zero (0-100)
-   `start_followup_date` date of first purchase of either AP or bactrim

Time-dependent information:

-   `bactrim` Purchase dates of bactrim
-   `AP` Purchase dates of AP
-   `karnofsky` date where Karnofsky score had dropped by at least one unit
-   `PCP` date where a pneumocystis pneumonia event was recorded

Outcome information: each person has either a date of death or a date
of last followup (censored date).


## Initialize an `rtmle` object and add the data

For the analysis we use 9 time intervals. 

```{r   }
x <- rtmle_init(intervals = 9,
                name_id = "id",
                name_outcome = "Death",
                name_competing = NULL,
                name_censoring = "Censored",
                censored_label = "censored")
x <- add_long_data(x,
                   outcome_data=register_data$death_data,
                   censored_data=register_data$censored_data,
                   competing_data=NULL,
                   timevar_data=register_data$timevar_data)
x <- add_baseline_data(x,data=register_data$baseline_data)
x$data$baseline_data
x$long_data 
```


## Discretize the continuous time data

To prepare the LTMLE analysis we discretize the time scale into 9
intervals, where each interval is 4 months long, starting from the
individual date where the followup starts, `start_followup_date`.
Then we map the long data to wide format, still keeping the
time-varying data in separate data.frames but readily combining the
outcome and the censored data into a single data.frame.

```{r   }
intervals <- seq(0,36*30.45,4*30.45)
x <- long_to_wide(x,
                  intervals = intervals,
                  fun = function(x){1 * (sum(x) > 0)},
                  start_followup_date = "start_followup_date")
x$data$timevar_data
x$data$outcome_data 
```


## Prepare data for analysis

The next step is to prepare a single wide format dataset where the
variables are sorted in time (from left to right):

```{r   }
x <- prepare_data(x)
x$prepared_data 
```

Before we continue it is useful to inspect the prepared data. To do
this, answer the following questions based on the prepared data:

1.  How many people initiate bactrim and how many initiate AP?
2.  By the end of the first interval, how many people had a PCP event, died or were censored?
3.  In the 3rd interval, how many people use bactrim, how many AP, how many both, and how many none?


## Formulate the protocols

We are now ready to formulate the target parameter:

```{r   }
x <- protocol(x,
              name = "Always_bactrim",
              intervention = data.frame("bactrim" = factor(rep("1",length(intervals)-1),levels = c("0","1")),"AP" = factor(rep("0",length(intervals)-1),levels = c("0","1"))))
x <- protocol(x,
              name = "Always_AP",
              intervention = data.frame("bactrim" = factor(rep("0",length(intervals)-1),levels = c("0","1")),"AP" = factor(rep("1",length(intervals)-1),levels = c("0","1"))))
x <- target(x,name = "Outcome_risk",
            estimator = "tmle",
            protocols = c("Always_bactrim","Always_AP"))
x$protocols
x$targets 
```


## Specify regression models

Finally, before running the analysis we need to specify which
variables should enter in the regression models. Note that we declare
the variable 'AP<sub>0</sub>' as constant because it is exactly equal to
'1-bactrim<sub>0</sub>' and hence would confuse the analyses. Note also that we
let the regression models only depend on the latest value of the
change in Karnofsky score and the occurrence of a PCP event.

```{r   }
x <- model_formula(x,exclude_variables = "AP_0",Markov = c("karnofsky","PCP"))
x$names$name_constant_variables <- c("AP_0",x$names$name_constant_variables)
x$models 
```


## Run the analysis

```{r   }
x <- run_rtmle(x,learner = "learn_glmnet")
summary(x) 
```


## Interprete the results

YOUR TEXT HERE


# Code for practicals IV

We continue to work with the register data of practicals III.  Now we
analyse the risk of a PCP event where death is a competing risk.


## Initialize an `rtmle` object and add the data

```{r   }
x <- rtmle_init(intervals = 9,name_id = "id",name_outcome = "PCP",name_competing = "Dead",name_censoring = "Censored",censored_label = "censored")
PCP_data <- register_data$timevar_data$PCP
setkey(PCP_data,id,date)
# restrict to first event
PCP_data <- PCP_data[PCP_data[,.I[1],by = "id"]$V1]
x <- add_long_data(x,outcome_data=PCP_data,competing_data = register_data$death_data,censored_data=register_data$censored_data,timevar_data=register_data$timevar_data[-1])
x <- add_baseline_data(x,data=register_data$baseline_data) 
```


## Discretize the continuous time data

```{r   }
x <- long_to_wide(x,
                  intervals = intervals,
                  fun = function(x){1 * (sum(x) > 0)},start_followup_date = "start_followup_date") 
```


## Prepare data for analysis

```{r   }
x <- prepare_data(x) 
```


## Formulate the protocols

```{r   }
x <- protocol(x,
              name = "Always_bactrim",
              intervention = data.frame("bactrim" = factor(rep("1",length(intervals)-1),levels = c("0","1")),"AP" = factor(rep("0",length(intervals)-1),levels = c("0","1"))))
x <- protocol(x,
              name = "Always_AP",
              intervention = data.frame("bactrim" = factor(rep("0",length(intervals)-1),levels = c("0","1")),"AP" = factor(rep("1",length(intervals)-1),levels = c("0","1"))))
x <- target(x,
            name = "Outcome_risk",
            estimator = "tmle",
            protocols = c("Always_bactrim","Always_AP")) 
```


## Specify regression models

```{r   }
x <- model_formula(x,
                   exclude_variables = "AP_0",
                   Markov = c("karnofsky"))
x$names$name_constant_variables <- c("AP_0",x$names$name_constant_variables) 
```


## Run the analysis

```{r   }
x <- run_rtmle(x,learner = "learn_glmnet")
summary(x)[,.(Protocol,Target_parameter,Estimate,P_value)] 
```


## Interprete the results

YOUR TEXT HERE

